Hamiltonian mechanics is conservation of information entropy

### 2019-05-22 Articles/HamiltonianMechanicsIsConservationOfEntropy/article.tex
Revision submitted to BJPS
Submitted to Studies
Adding section on physicality of information entropy

\documentclass[11pt]{article}

@@

\title{Hamiltonian mechanics is conservation of information entropy}
%\author{Gabriele Carcassi, Christine A. Aidala}

%\ifjournal
%	% As the page margins are so large, the full title does not fit
%	\titlerunning{From physical principles to classical and quantum particle mechanics}
%\fi

%\date{\today}

@@

This insight helps to clearly characterize what Hamiltonian mechanics describes, providing a clear physical meaning to the mathematical structures used. This should be at least of interest to that part of the philosophy community interested in the foundations of classical mechanics. Second because information entropy has a deep connection to the foundations of thermodynamics, therefore the strong link we establish here between information entropy and Hamiltonian mechanics should be of interest to an even wider audience.

@@

%Now, the fact that we can use a tangent bundle to track our particle states does \emph{not} mean that conjugate momentum is a one form, in the sense of a map from a vector to a number. The state space is just a manifold, a set of points identified by $n$ continuous variable, and the structure it has comes from how units transform, not because the variables identify separate objects. Unfortunately mathematical structures strip out the units, the knowledge of the double role of coordinates is lost and then a different mathematical concept is used, one that has the same mathematical features of our particle state space but an altogether different meaning and construction. Changing state variables from position and momentum to position and velocity is not, as one may mistakenly understand from the math, a change of objects from a cotangent bundle to a tangent bundle: it is relabeling the same states much like changing spatial coordinates relabels positions. This change of variables, though, is not in general a canonical transformation which means the density will not correspond to the invariant one. But it is still a valid change of state variables and in fact, giving that velocity is gauge invariant, a very useful one. It allows to write Poisson brackets and find invariants of motion that are readily physically meaningful~\cite{Holten}.

%We can summ

% In fact, if $(q^1, p_1)$ and $(q^2, p_2)$ are two degrees of freedom and $\rho_1(q^1, p_1)$ and $\rho_2(q^2, p_2)$ are the respective marginal distributions, we have now four ways of characterizing that they are independent.
%These characterizations are equivalent because they repre

%This means that relationships between the entropy of the total and marginal distributions are preserved. For example, saying that the marginal distributions $\rho_1(q^1, p_1)$ and $\rho_2(q^2, p_2)$ are independent is equivalent to saying that the entropy of the joint distribution is equal to the sum of the entropy of the marginals: $I(\rho(q^1,q^2, p_1, p_2)) = I(\rho_1(q^1, p_1)) +I(\rho_2(q^2, p_2))$.\footnote{This is also equivalent to saying that the joint distribution is the product of the marginals $\rho(q^1,q^2, p_1, p_2) = \rho_1(q^1, p_1)\rho_2(q^2, p_2)$.} That is, giving both values is equivalent to giving one and then the other. This means we can think of these relationships and quantities as real physical properties assigned to the degree of freedom and not just to the coordinates.

%\begin{description}
%	\item \emph{Coordinate independence}. The unit of $q^1$ can be chosen independently of $q^2$.
%	\item \emph{State count}. If $dq^1 dp_1$ and $dq^2 dp_2$ represent how many possible states the count of states in each degree of freedom then the total count across is the product 
%	\item \emph{Statistical independence}. The joint distribution is the product of the marginals: $\rho(q^1,q^2, p_1, p_2) = \rho_1(q^1, p_1)\rho_2(q^2, p_2)$.
%	\item \emph{Information independence}. The information entropy of the joint distribution is the sum of the information entropy of the marginals: $I(\rho(q^1,q^2, p_1, p_2)) = I(\rho_1(q^1, p_1)) +I(\rho_2(q^2, p_2))$. Intuitively this means that giving the value of all state variables is equivalent to giving the values of both degree of freedom one after the other.
%\end{description}
%These characterizations are equivalent: you can't have one without the others.

@@

@@

As we saw, Hamiltonian dynamics not only conserves the energy of individual particles, but it also conserves the information entropy of the distribution as a whole. Intuitively, we can see that there is some link between the spread of a distribution and the number of bits required to identify an element. For example, if we have two uniform distributions with different ranges, then Alice will need to give Bob more information to identify the value from the bigger range. So we ask: what is the distribution that minimizes the spread given a certain amount of information entropy? If $I$ is the set amount of information entropy and $\sigma_q^2 \sigma_p^2 \equiv \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$ is the spread, we can use Lagrange multipliers to find out.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1) \\ &+ \lambda_2(- \int \rho \ln \rho \, dqdp - I)\\ 
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 + \\ &\lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I = &\ln (2\pi\sigma_q\sigma_p) + 1
\end{align*}
The distribution that minimizes the spread, then, is the product of two independent Gaussians. As the entropy is conserved during Hamiltonian evolution the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I - 1) / 2 \pi 
\end{align*}
Hamiltonian mechanics is the deterministic and reversible evolution for a distribution and its elements, and therefore it cannot reduce the spread of the distribution indefinitely. If it did, it would start concentrating the density, meaning that multiple states would be mapped to a single state and the dynamics would not be reversible. In this form $I$ is not a fundamental constant but a parameter that depends upon the initial condition. While this is already striking,~\cite{AoPPhy1} shows a simple argument to turn this into something more.

@@

We can then decide to measure volumes in phase space in terms of this minimum spread, which means setting the value of $\hbar$ (or $h = 2 \pi \hbar$) according to the following relationship:
\begin{align*}
\frac{\hbar}{2} &= \frac{h}{4\pi} \equiv \frac{e^{I_0 - 1}}{2 \pi} =  \frac{e^{I_0}}{2 e \pi} \leq \sigma_q\sigma_p
\end{align*}
 This is what~\cite{AoPPhy1} refers to as the classical uncertainty principle.

@@

%These results represent one set of insights from our larger project, Assumptions of Physics, that aims to rederive the standard laws from a handful of physical starting points.

>>>>

\documentclass[11pt]{elsarticle}

@@

\title{Hamiltonian mechanics is conservation of information entropy}

@@

This insight helps to clearly characterize what Hamiltonian mechanics describes, providing a clear physical meaning to the mathematical structures used. This should be at least of interest to that part of the philosophy community interested in the foundations of classical mechanics. Secondly, information entropy seems to have a connection to the foundations of thermodynamics and statistical mechanics, in the sense that it is used successfully in physics practice, even on systems far from equilibrium (cite the 2 papers that Horowitz gave us). Therefore the strong link we establish here between information entropy and Hamiltonian mechanics should be of interest to an even wider audience.

@@

@@

\subsection*{On the physicality of information entropy}

We want to stress, again, that in general information entropy cannot be taken as an objective physical quantity. It depends on the choice of symbols one uses for encoding: using letters of the alphabet or words in the English language will yield a different information entropy for the same book. It will depend on the choice of ensemble: in game theory one uses credence distributions that are subjective objects and therefore the information entropy will be different for different actors. The nature of the information entropy, then, depends on the nature of the elements, the ensemble and the values over which the distribution is defined.

Moreover, we should note that our construction is not the one typically used in classical statistical mechanics. In that case, the elements are points across $N$ copies of phase space (one set of position and momentum for each particle) and the distribution over them is a probability distribution. In this case one has to be careful and show how the distribution is uniquely defined by objective physical entities. If one interprets said distribution as credence, like (cite Janes), then one has obvious problems.

Our case, then, is by construction very particular. The ensemble is the state of a physical system. The elements are the physical subdivisions of said system, the particles. The set of values over which the distribution is defined is the state space of the parts. The densities relate what fraction of the whole can be found in a particular particle state. Densities and areas in terms of $q^i$ and $k_i$ are invariant pure numbers, independent of the choice of unit. No ``partitioning" or ``averaging" of phase space is invoked. The information entropy depends on no other external factors: it is an invariant state variable of the ensemble. It is a physical property of the system.

Our statistical distribution does not correspond to the probability distribution of statistical mechanics. Conceptually, it corresponds to a point across the $N$ copies of phase space, a microstate, in the following way: pick a very narrow unitary distribution $\delta(q,p)$; make a copy for each particle so that it is centered around its position and momentum; sum them all.\footnote{Note that if we permute the particles we obtain the same distribution, so this repackaging does not suffer from overcounting.} That is:
\begin{equation}
\rho(q,p) = \sum\limits_{a=1}^N \delta(q - q_a, p - p_a)
\end{equation}
To form a complete link to statistical mechanics we would have to define what the probability distribution is. To form a complete connection to thermodynamics we would have to introduce a link to thermodynamic entropy. This will be outside of the scope of this work. At this point we simply want to show how information entropy applied to statistical distributions over phase space is indeed a physical quantity that is deeply linked to the notion of deterministic and reversible evolution and to Hamiltonian mechanics.

@@

As we saw, Hamiltonian dynamics not only conserves the energy of individual particles, but it also conserves the information entropy of the distribution as a whole. Intuitively, we can see that there is some link between the spread of a distribution and the number of bits required to identify an element. For example, if we have two uniform distributions with different ranges, then Alice will need to give Bob more information to identify the value from the bigger range. So we ask: what is the distribution that minimizes the spread given a certain amount of information entropy? If $I$ is the set amount of information entropy and $\sigma_q^2 \sigma_p^2 \equiv \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$ is the spread, we can use Lagrange multipliers to find out.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1) \\ &+ \lambda_2(- \int \rho \ln \rho \, dqdp - I)\\ 
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 + \\ &\lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I = &\ln (2\pi e\sigma_q\sigma_p)
\end{align*}
The distribution that minimizes the spread, then, is the product of two independent Gaussians. As the entropy is conserved during Hamiltonian evolution the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I) / 2 \pi e 
\end{align*}
Hamiltonian mechanics is the deterministic and reversible evolution for a distribution and its elements, and therefore it cannot reduce the spread of the distribution indefinitely. If it did, it would start concentrating the density, meaning that multiple states would be mapped to a single state and the dynamics would not be reversible. In this form $I$ is not a fundamental constant but a parameter that depends upon the initial condition. While this is already striking,~\cite{AoPPhy1} shows a simple argument to turn this into something more.

@@

We can then decide to measure volumes in phase space in terms of this minimum spread, which means setting the value of $\hbar$ (or $h = 2 \pi \hbar$) according to the following relationship:
\begin{align*}
\frac{\hbar}{2} &= \frac{h}{4\pi} \equiv  \frac{e^{I_0}}{2 e \pi} \leq \sigma_q\sigma_p
\end{align*}
 This is what~\cite{AoPPhy1} refers to as the classical uncertainty principle.

@@
